{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-20 thesis/education.tex
impact on grades (photo of ze studence)

I evaluated Landslide's impact on student submission quality in two ways.
First, I analyzed the overall P2 grades of CMU 15-410 students between Landslide users and non-users,
to see if Landslide helped them submit overall better implementations,
as well as their later overall P3 (kernel project) grades,
to see if Landslide helped them internalize any concurrency debugging skills
that would help write better kernels even without Landslide's aid.
% This experiment produced inconclusive or negative results. % say this?
%
Second, I studied
several % how many? 3
individual concurrency bugs that I thought Landslide was likely to detect,
all already well-known by 15-410 course staff,
to see if using Landslide correlated with submitting projects absent those bugs.
% results..?

@@

Figure~\ref{Fig:photo-of-ze-studence} shows some resulce

\begin{figure}[t]
	\includegraphics[width=\textwidth]{photo-of-ze-studence.pdf}
	\caption{sadfasdfa}
	\label{fig:photo-of-ze-studence}
\end{figure}

@@

Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
I selected 3 bugs to study in depth to ascertain
whether Landslide played an instrumental difference
in helping the students ultimately submit respecitvely correct implementations.
To avoid bias of picking too obscure and/or trivial bugs
that Landslide alone might find but even course staff would not expect students to solve,
I chose only bugs which had substantial penalties in the grading rubric
(guided, as well, by my own intuition as a former TA).
I also counted students who used Landslide in general but whose usage snapshots showed they did not run the test case in question as non-users.
Table~\ref{tab:eval-common-bugs} presents the results.
% two-tailed btw, the better (more conservative) way,
% tbh one tailed might even make sense here, but it doesn't change anything wrt the cutoffs anyway, so
The $p$ values are calculated in R \cite{r-lang} using Fisher's exact test \cite{fishers-exact-test},
treating each row as an independent 2x2 contingency table,
and dividing the significance cutoff by the number of bugs to conservatively account for multiplicity \cite{xkcd-jellybeans}.

@@

{\bf Thread exit/join.}
%During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
After a thread finishes exiting, the memory allocated for its stack must be reclaimed for other uses.
A common student pitfall is to allow a {\tt thr\_join()}ing thread to free said stack space
while {\tt thr\_exit()} is still executing userspace C code (which typically accesses the stack),
or even for {\tt thr\_exit()} to free it itself.
In the former case, threads must interleave specifically to exhibit a use-after-free;
in the latter case, the use-after-free will be deterministic (i.e., present in all interleavings).
Other than Landslide, however,
the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access.
This means that a subsequent {\tt thr\_create()} invocation would need to race to recycle the memory for a new thread stack
and conflict with the old exiting thread
before any problem could be detected.
% TODO: mark which of the tej bugs were deterministic vs race
The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug.
%, and
%the official rubric penalties for it are
%{\tt :exit\_stack\_minor -2} and
%{\tt :exit\_stack\_race -4} (for the former case),
%and
%{\tt :uses\_freed\_stack -4} (for the latter).
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.
%
This bug is especially remarkable in that not a single student who did not use Landslide these semesters
submitted a correct implementation.

{\bf Mutex.}
The Landslide-friendly {\tt mutex\_test} checks for the possibility of two contending threads
accessing a mutex-protected critical section simultaneously.
It includes one thread repeating once the lock, unlock cycle
so as to check for unlock/lock races as well as lock/lock interactions,
and, as mentioned previously (\sect{\ref{sec:education-pebbles-tests}}),
checks for data race access pairs inside the mutex implementation itself.
As students have free rein to design their mutex internals,
this refers to the general class of mutex bugs
in which any number of things can go wrong, depending on the implementation,
leading to mutual exclusion (or otherwise assertion) failure.
Whether each student's submitted P2 solved or suffered from such bugs
was determined by checking their grade files for any mutex-related penalties (assessed by the TAs),
then me double-checking their implementations by hand to confirm.

@@

{\bf Paraguay.}
Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
can spuriously wake up early from the second {\tt cond\_wait()}.
The precise reasoning why many na\"ive student implementations are susceptible to this bug,
as well as the 3 major ways of fixing it,
are one of 15-410 staff's closely-guarded secrets (\sect{\ref{sec:410-secrecy}}).
Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
which biases thread scheduling towards the interleaving required to exhibit the bug
(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
Hence, despite being a subtle concurrency bug,
the official course test suite is likely to expose it,
so comparing how many Landslide users and non-users submitted this bug in particular
speaks more to the impact of Landslide's preemption traces
in helping students diagnose a bug that a ``stress'' test could already find.
%
{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
%, and
%the official rubric penalty for it is
%{\tt :paraguay -2}.
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

{\bf R/W-lock downgrade.}
In addition to the standard R/W-lock interface,
P2 requires students to implement {\tt rwlock\_downgrade()}:
called with the lock held in write mode,
the caller adopts the reader role instead,
allowing other waiting reader threads to proceed simultaneously,
all while allowing no waiting writers to access in between.
%
The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
checks that readers are allowed simultaneous access after a downgrade with no possibility for deadlock.
Rather than one specific bug, this refers to any of several failures that can arise during a downgrade.
%, and
%the official rubric penalties related to this bug are
%{\tt :downgrade\_no\_signal -1},
%%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
%{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
%{\tt :downgrade\_\allowbreak{}lies -3}, and
%{\tt :downgrade\_deadlock -4}.
Whether each student's submitted P2 solved or suffered from these bugs
was determined by checking their grade files for any downgrade-related penalties
(which were assessed by the TAs as a result of their manual inspection, rather than mine).

>>>>

I evaluated Landslide's impact on student submission quality in two ways.
First, I analyzed CMU 15-410 students' overall project grades between Landslide users and non-users
to see if Landslide helped them submit overall better implementations.
Second, I studied
several % how many? 3
individual concurrency bugs that I thought Landslide was likely to detect,
all already well-known by 15-410 course staff,
to see if using Landslide correlated with submitting projects absent those bugs.
% results..?

@@

Figure~\ref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
between Fall 2013 and Spring 2018,
grouped by whether or not students used Landslide during P2.
%Landslide users are, of course, the experimental group.
The control group is further distinguished by whether Landslide was offered and the student delined,
or whether the study had not yet begun that semester.
Intuition suggests students in the former group would be more likely
to be struggling too much to have free time to volunteer in the first place,
so any comparison between them and Landslide users is vulnerable to selection bias.
The latter control group, in which students did not have a choice, mitigates such bias,
although itself may be vulnerable to other confounding factors such as grading criteria varying across semesters.
%
In addition to P2, I also show the subsequent P3 (kernel project) grades,
likewise broken down by who used Landslide previously during P2%
\footnote{The P3 distributions are slightly smaller than those from P2,
as some students dropped the class in between.
A few new project groups also tend to appear during P3 as partnerships break up or re-form,
but as these cannot be meaningfully classified as Landslide users or non-users,
I omit them here.
}.
Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
one explanation could be that Landslide helped students
internalize new concurrency programming and/or debugging skills
that would help them write better kernels even without Landslide's aid.
%

% TODO: contingency table for drop rate? but.... need to get studence names to make sure they didn't just solbo, or w/e

% colors for these graphs in case i need them again later
% #720000
% #FF8080
% #804700
% #FFC680
% #807700
% #FFF780
% #004111
% #ABFFC0
% #002680
% #ABC3FF
% #3B0144
% #F4ABFF
\begin{figure}[p]
	\begin{center}
	\begin{tabular}{c}
		\includegraphics[width=0.96\textwidth]{photo-of-ze-studence.pdf}
		\\
		(a) Visualized as a box plot.
		\\
		\\
		\includegraphics[width=0.96\textwidth]{photo-of-ze-studence-cdf.pdf}
		\\
		(b) Visualized as a CDF.
	\end{tabular}
	\end{center}
	\caption{Distribution of project grades among Landslide users and non-users.
	Study participants (S'15 to S'18) are the experimental group;
	non-participants (S'15 to S'18)
	and students from semesters predating the study (F'13-F14) are the control groups.
	Compare Landslide users to non-users for within-semester differences,
	or users to pre-study students to mitigate selection bias.}
	\label{fig:photo-of-ze-studence}
\end{figure}

%We see in Figure~\ref{fig:photo-of-ze-studence} that, indeed,
{\bf Results.}
Overall,
Landslide users did indeed receive
slightly better grades overall on their P2 submissions than non-users from the same semesters,
although comparing to the second control group lessens that difference somewhat.
The difference is also less pronounced in both comparisons among P3 grades
(unsurprisingly, as learning lasting lessons about concurrency should be harder than fixing bugs case-by-case),
although the experimental group still maintains a tiny lead.

{\bf Statistical significance.}
Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess statistical significance,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two CDFs,
testing if their samples are likely to arise from the same underlying, though unspecified, distribution%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between CDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}.
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons (see next section)
because each one is testing a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses (such as ``more skilled students were more likely to sign up to begin with'').

Ultimately, only the same-semester comparisons were significant
However, the distribution difference being smaller in P3 than in P2
% TODO: is there a good way to treat this more formally?
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias.
% let me have this exclamation mark. please
That aside, ``Landslide helps students submit better P2s'' is the next most likely hypothesis I can think of!

% P2 vs opt out
% ad.test(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(38,42,47,50,52,53,55,55,56,56,57,57,58,58,58,59,60,60,61,62,62,63,65,66,67,68,68,68,69,70,70,70,71,71,71,71,72,72,72,73,73,73,74,74,76,76,76,77,77,78,78,78,78,78,79,79,79,80,82,82,83,83,83,84,84,84,85,86,86,88,90,95,96,96))
%               AD  T.AD  asympt. P-value
% version 1: 2.611 2.139          0.04263
% version 2: 2.760 2.334          0.03576

% P2 vs pre study
% ad.test(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(43,51,56,58,59,59,60,60,60,60,61,61,61,61,61,62,63,65,65,67,68,69,69,69,69,69,70,70,71,71,72,72,72,73,73,73,73,73,74,74,74,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,79,79,80,80,80,80,80,80,81,81,82,82,82,82,83,83,83,83,85,85,85,86,86,86,88,88,90,92,93))
%                AD    T.AD  asympt. P-value
% version 1: 0.6248 -0.4977           0.6275
% version 2: 0.5790 -0.5581           0.6706

% P3 vs opt out
% ad.test(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(45,58,66,70,72,72,72,72,74,74,74,75,75,75,77,77,77,78,78,78,79,79,79,79,79,79,80,80,81,81,81,81,81,82,82,83,83,83,83,83,84,84,85,85,86,86,86,86,86,86,87,87,89,89,89,89,90,91,91,91,93,93,93))
%               AD  T.AD  asympt. P-value
% version 1: 2.645 2.186          0.04085
% version 2: 2.800 2.398          0.03376

% P3 vs pre-study
% ad.test(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(60,62,63,64,65,68,69,72,72,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,80,80,80,81,81,81,81,81,81,81,81,81,82,82,82,82,82,83,83,83,84,84,84,84,84,84,84,85,86,86,87,87,88,88,88,88,88,88,88,89,89,89,89,89,90,90,90,91,92,92,92,93,94,95))
%               AD   T.AD  asympt. P-value
% version 1: 1.552 0.7322           0.1633
% version 2: 1.640 0.8473           0.1454

\begin{figure}[t]
	\begin{center}
		\small
	\begin{tabular}{cr|r||r|r|r|r||cc}
					&	&	&	&	& \multicolumn{2}{c}{Significance}
		\multicolumn{2}{c|}{Population}
					& $N$	& min	& mean	& max	& stddev & $p$ vs users	& cutoff\\
		\hline
			& users		& 103	& 42	& 74.8	& 95	& 10.3	& --	& --	\\
		P2	& non-users	& 74	& 38	& 70.9	& 96	& 12.5	& 0.036	& 0.05	\\
			& pre-study	& 86	& 43	& 73.7	& 93	& 9.7	& 0.671	& 0.05	\\
			\hline
			& users		& 98	& 65	& 83.6	& 95	& 6.2	& --	& --	\\
		P3	& non-users	& 63	& 45	& 80.7	& 93	& 8.2	& 0.034	& 0.05	\\
			& pre-study	& 79	& 60	& 81.6	& 95	& 7.6	& 0.145	& 0.05	\\
	\end{tabular}
	\end{center}
	\caption{Detailed statistics from student grade distributions.}
	\label{tab:table-of-ze-studence}
\end{figure}

@@

Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
I selected 3 bugs to study in depth to ascertain
whether Landslide played an instrumental difference
in helping the students ultimately submit respecitvely correct implementations.
To avoid bias of picking too obscure and/or trivial bugs
that Landslide alone might find but even course staff would not expect students to solve,
I chose only bugs which had substantial penalties in the grading rubric
(guided, as well, by my own intuition as a former TA).
I also counted students who used Landslide in general but whose usage snapshots showed they did not run the test case in question as non-users.
Table~\ref{tab:eval-common-bugs} presents the results.

@@

Each of the four bugs studied is described in detail as follows.

\begin{enumerate}
\item
{\bf Thread exit/join.}
%During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
After a thread finishes exiting, the memory allocated for its stack must be reclaimed for other uses.
A common student pitfall is to allow a {\tt thr\_join()}ing thread to free said stack space
while {\tt thr\_exit()} is still executing userspace C code (which typically accesses the stack),
or even for {\tt thr\_exit()} to free it itself.
In the former case, threads must interleave specifically to exhibit a use-after-free;
in the latter case, the use-after-free will be deterministic (i.e., present in all interleavings).
Other than Landslide, however,
the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access.
This means that a subsequent {\tt thr\_create()} invocation would need to race to recycle the memory for a new thread stack
and conflict with the old exiting thread
before any problem could be detected.
% TODO: mark which of the tej bugs were deterministic vs race
The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug.
%, and
%the official rubric penalties for it are
%{\tt :exit\_stack\_minor -2} and
%{\tt :exit\_stack\_race -4} (for the former case),
%and
%{\tt :uses\_freed\_stack -4} (for the latter).
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.
%
This bug is especially remarkable in that not a single student who did not use Landslide these semesters
submitted a correct implementation.

\item
{\bf Mutex.}
The Landslide-friendly {\tt mutex\_test} checks for the possibility of two contending threads
accessing a mutex-protected critical section simultaneously.
It includes one thread repeating once the lock, unlock cycle
so as to check for unlock/lock races as well as lock/lock interactions,
and, as mentioned previously (\sect{\ref{sec:education-pebbles-tests}}),
checks for data race access pairs inside the mutex implementation itself.
As students have free rein to design their mutex internals,
this refers to the general class of mutex bugs
in which any number of things can go wrong, depending on the implementation,
leading to mutual exclusion (or otherwise assertion) failure.
Whether each student's submitted P2 solved or suffered from such bugs
was determined by checking their grade files for any mutex-related penalties (assessed by the TAs),
then me double-checking their implementations by hand to confirm.

@@

\item
{\bf Paraguay.}
Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
can spuriously wake up early from the second {\tt cond\_wait()}.
The precise reasoning why many na\"ive student implementations are susceptible to this bug,
as well as the 3 major ways of fixing it,
are one of 15-410 staff's closely-guarded secrets (\sect{\ref{sec:410-secrecy}}).
Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
which biases thread scheduling towards the interleaving required to exhibit the bug
(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
Hence, despite being a subtle concurrency bug,
the official course test suite is likely to expose it,
so comparing how many Landslide users and non-users submitted this bug in particular
speaks more to the impact of Landslide's preemption traces
in helping students diagnose a bug that a ``stress'' test could already find.
%
{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
%, and
%the official rubric penalty for it is
%{\tt :paraguay -2}.
Whether each student's submitted P2 solved or suffered from this bug
was determined by me personally inspecting their code.

\item
{\bf R/W-lock downgrade.}
In addition to the standard R/W-lock interface,
P2 requires students to implement {\tt rwlock\_downgrade()}:
called with the lock held in write mode,
the caller adopts the reader role instead,
allowing other waiting reader threads to proceed simultaneously,
all while allowing no waiting writers to access in between.
%
The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
checks that readers are allowed simultaneous access after a downgrade with no possibility for deadlock.
Rather than one specific bug, this refers to any of several failures that can arise during a downgrade.
%, and
%the official rubric penalties related to this bug are
%{\tt :downgrade\_no\_signal -1},
%%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
%{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
%{\tt :downgrade\_\allowbreak{}lies -3}, and
%{\tt :downgrade\_deadlock -4}.
Whether each student's submitted P2 solved or suffered from these bugs
was determined by checking their grade files for any downgrade-related penalties
(which were assessed by the TAs as a result of their manual inspection, rather than mine).
\end{enumerate}

{\bf Statistical significance.}
% two-tailed btw, the better (more conservative) way,
% tbh one tailed might even make sense here, but it doesn't change anything wrt the cutoffs anyway, so
The $p$ values
in Table~\ref{tab:eval-common-bugs}
are calculated in R \cite{r-lang} using Fisher's exact test \cite{fishers-exact-test},
treating each row as an independent 2x2 contingency table.
I divide 0.05, the standard significance cutoff,
by the number of bugs to conservatively account for multiplicity \cite{xkcd-jellybeans}
(not that it matters).
% TODO write more
